### 2022-08-24 main/chapter_searches.tex
add gaussian stuff

\paragraph{Probability assignments.}
Probability algebra manipulates probabilities, but does not state how they
should be initially assigned.
In general, such probability assignments are free and unconstrained, and might
informed by calibration data or loosely motivated opinions.
But there are examples of unambiguously derived that are commonly used in
High-Energy Physics, notably the Poisson and Gaussian distributions.
To avoid turning this thesis into a textbook, we only sketch the derivations
of these, and refer to real textbooks for their detailed
derivations~\cite{jaynes2003probability}, both of which use the exponential
identity that
\begin{equation}
\label{eqn:searches_exponential}
\exp(x) = \displaystyle \lim_{N \to \infty} (1 - \frac{x}{N})^N
.
\end{equation}

The Large Hadron Collider collides $40$~million proton bunches per second,
but \atlas' trigger system accepts only only around $1000$ of
those~\cite{atlas2020trigger}.
This is a Binomial scenario with a large fixed number $N$ of trials
(bunch collisions)
and a small probability of acceptance (triggering),
due to a notable collision event.
The Binomial distribution for $n$ triggered events is well known as
\begin{equation}
\mathrm{Binomial}(n\mid N, p) = p^n (1 - p)^{N - n} \times \binom{N}{n}
,
\end{equation}
which is a product of the probabilities to the sequence of $N$ accept/reject
data, summed over the permutations that could have led to the total $n$.
With our extremely small trigger rate, we are interested in the limit of small
$p$ and large $N$ which result in a mean acceptance number of $\lambda = Np$.
In the large $N$ limit, this tends towards the Poisson
distribution~\cite{jaynes2003probability}
\begin{equation}
\label{eqn:searches_poisson}
\mathrm{Poisson}(n\mid \lambda) = \frac{\lambda^n}{n!}\exp(-\lambda)
,
\end{equation}
which is used ubiquitously in the analysis of our histograms.

The Gaussian distribution has many justifications:
as a practical approximation it permits analysis in linear algebra,
as the unique distribution is uniquely spherically symmetric and factors into
independent dimensions~\cite{
jaynes2003probability,
herschel1850normal,
maxwell1860normal,
muller1959note,
marsaglia1972normal
},
and as the maximum entropy~\cite{PhysRev.106.620} distribution for a
known mean and variance.
\TODO{}
Succinctly stated by Jaynes that
``any function with a single rounded maximum, raised to a higher and higher
power, goes into a Gaussian function''~\cite{jaynes2003probability}.
\begin{equation}
\label{eqn:searches_gaussian}
\mathrm{Gaussian}(x\mid \mu, \sigma) =
\frac{1}{\sqrt{2\pi \sigma^2}}
\exp\left(-\frac{1}{2}\frac{(x - \mu)^2}{\sigma^2}\right)
\end{equation}
This is not the famous Central Limit Theorem, which is related to the
distribution of sums, but can be used to derive it: the distribution of a sum
is a convolution integral, which becomes a product in Fourier-transformed
space, and that product can converge to a Gaussian function by this rule.
Jaynes gives a beautiful alternative proof of the
Central Limit Theorem~\cite{jaynes2003probability}
(Section ``7.16 The central limit theorem'').

@@

% connection to frequentist section
Bayes' theorem is an infamous and necessary result, and it is not seriously
disputed. What has been disputed is whether probability can be applied to
logic, or any application other than the random happenings of events.

>>>>

\paragraph{Probability assignments.}
Probability algebra manipulates probabilities, but does not state how they
should be initially assigned.
In general, such probability assignments are free and unconstrained, and might
informed by calibration data or loosely motivated opinions.
But there are examples of unambiguously derived that are commonly used in
High-Energy Physics, notably the Poisson and Gaussian distributions.
To avoid turning this thesis into a textbook, we only sketch the derivations
of these, and refer to real textbooks for their detailed
derivations~\cite{jaynes2003probability}, both of which use the exponential
identity that
\begin{equation}
\label{eqn:searches_exponential}
\exp(x) = \displaystyle \lim_{N \to \infty}
\left(1 - \frac{x}{N}\right)^N
.
\end{equation}

The Large Hadron Collider collides $40$~million proton bunches per second,
but \atlas' trigger system accepts only only around $1000$ of
those~\cite{atlas2020trigger}.
This means we have scenario with a large fixed number $N$ of trials
(bunch collisions)
and a small probability of binary acceptance (triggering),
due to a notable collision event.
The summed total number $n$ of accepted events is therefore described by the
Binomial distribution for $n$;
\begin{equation}
\mathrm{Binomial}(n\mid N, p) = p^n (1 - p)^{N - n} \times \binom{N}{n}
,
\end{equation}
which is a product of the probabilities (produce rule) in the sequence of $N$
accept/reject decisions, summed (sum rule) over all of that sequence
permutations that could have led to the total $n$.
With our extremely small trigger rate, we are interested in the limit of small
$p$ and large $N$ which result in a mean acceptance number of $\lambda = Np$.
Using the exponential identity of Equation~\ref{eqn:searches_exponential}, one
can show that this limit tends towards the Poisson
distribution~\cite{jaynes2003probability};
\begin{equation}
\label{eqn:searches_poisson}
\mathrm{Poisson}(n\mid \lambda) = \frac{\lambda^n}{n!}\exp(-\lambda)
,
\end{equation}
which is used ubiquitously in the analysis of our histograms.

The Gaussian distribution has many justifications:
as the unique distribution that is spherically symmetric and factors linearly
into independent dimensions~\cite{
jaynes2003probability,
herschel1850normal,
maxwell1860normal,
muller1959note,
marsaglia1972normal
},
as a practical approximation that permits analysis in linear algebra,
and as the maximum entropy~\cite{PhysRev.106.620} distribution for a
known mean and variance.
This maximum entropy statement means that for a distribution with known mean
and variance in a given coordinates, its best Gaussian approximation matches
those moments.\footnote{%
By `best', this means minimal Kullback-Leibler divergence
\(
H(p\leftarrow q) = \int \log(p(x)/q(x)) \,\mathrm{d}p(x)
\)
from the Gaussian approximation $q$ to the target $p$ with known moments.
Such an approximation may not be \emph{good}, but it is the best in its class.%
}
As a limiting approximation, the Gaussian form arises from products of many
functions, such as combinations of many independent measurements that multiply
their likelihood functions.
At a maximum with $x=\mu$, the first derivative of a function vanishes and its
second derivative is negative, so this product looks something like
\begin{equation}
f(x)^N = \displaystyle \lim_{N \to \infty}
(1 - \kappa (x - \mu)^2)^N,
\end{equation}
which, again by Equation~\ref{eqn:searches_exponential}, goes into the Gaussian
form
\begin{equation}
\label{eqn:searches_gaussian}
\mathrm{Gaussian}(x\mid \mu, \sigma)
\,\mathrm{d}x =
\frac{1}{\sqrt{2\pi \sigma^2}}
\exp\!\left(-\frac{1}{2}\frac{(x - \mu)^2}{\sigma^2}\right)
\mathrm{d}x
\end{equation}
where the prefactor is for normalization to a probability distribution over
$x$, and we include the $\mathrm{d}x$ terms as reminders that it is a density
function in that coordinate.

Here, $\kappa$ and $\sigma$ both relate to the second derivative of the target
function; this relationship extends to multiple dimensions in the Hessian matrix
and its inverse, and allows one to estimate error bars from second derivatives
--- a practical tool that the methods discussed in this thesis employ.

This is not the famous Central Limit Theorem, which is related to the
distribution of sums.
But can be used to derive it: the distribution of a sum is a convolution, which
becomes a product in Fourier-transformed space, and that product can converge
to a Gaussian function by this same derivation if the mean and variance exist.
A beautiful alternative proof of the Central Limit Theorem is given by
Jaynes~\cite{jaynes2003probability}
(Section ``7.16 The central limit theorem''), and would be spoiled by a
reproduction here.

@@

\TODO{}
% connection to frequentist section
Bayes' theorem is an infamous and necessary result, and it is not seriously
disputed. What has been disputed is whether probability can be applied to
logic, or any application other than the random happenings of events.
